Hydra Battles and Cie

### 2021-01-15 doc/chapter-powers.tex
comments and examples about the Ackermann part (to continue)

\subsection{A common notation for multiplication}
\label{op-classes}
\index{Type Classes!Operational Type Classes}
\emph{Operational type classes}~\cite{BS2011}
allow us to define a common notation 
for multiplication in any algebraic structure. 
First, we associate a class to the notion of \emph{multiplication} 
on any type $A$.

@@

\subsection{The Monoid Type Class}
\index{Type Classes}
We are now ready to  give a definition of the \texttt{Monoid} class, using
\texttt{*} as an infix operator in scope \coqscope{M} for the monoid  multiplication.

@@

\subsection{Monoids and Equivalence Relations}
\index{Generalized rewriting}
\index{Equivalence relations}

@@

\index{Type Classes!Proper}
\label{EMonoid-def}

@@

\index{Coercions}

@@

In order to make possible proof by rewriting on expressions which contain
the exponentiation operator, we have to prove that, whenever \texttt{$x$ == $y$},
the equality \texttt{$x^n$ == $y^n$} holds for any exponent \texttt{$n$}. 
For this purpose, we use the \texttt{Proper} class of module
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Classes.Morphisms.html}{Coq.Classes.Morphisms}
\index{Type Classes!Proper}

\begin{Coqsrc}
Global Instance power_proper :
     Proper (equiv ==> eq ==> equiv) power.
(* Proof omitted *)
\end{Coqsrc}

@@

\subsection{Correctness Proofs by Reflection}
\label{reflection-section}
\index{Advanced proof techniques!Proof by reflection}
Instead of letting the tactic \texttt{rewrite} look for contexts in which
setoid rewriting is possible, we propose to use (deterministic) computations for
obtaining a ``canonical'' form for terms generated from a variable \texttt{x}
by contructors associated with monoid multiplication and neutral element.

@@

\subsubsection{About parametricity}
\index{Advanced proof techniques!Parametricity}
Let us now present another tactic for proving chain correctness,
in the tradition of works on \emph{parametricity} and its use for 
proving properties on programs.
Strachey~\cite{Strachey:2000:FCP:609150.609208}
explores the nature of \emph{parametric
polymorphism}: ``\emph{Polymorphic functions behave uniformly for all types}''
then Reynolds~\cite{REYNOLDS83} formalizes this notion through binary relations.
Wadler~\cite{Wadler1989}, then Cohen \emph{et al.}~\cite{Cohen2013}
use this relation for deriving
 theorems about functions that operate on parametric
polymorphic types.

@@

\subsubsection{Definition:} 
\index{Type Classes!Proper}

A continuation \texttt{k:Fkont A} is \emph{proper}
if, whenever \linebreak[3] \texttt{x == y} holds, the computations \texttt{k x} and 
\texttt{k y} are equivalent.

@@

The following dependently-typed functions will help us to specify  formally
any correct chain generator.
\index{Dependently Typed Functions}

>>>>

\subsection{A common notation for multiplication}
\label{op-classes}
\index{Coq!Type Classes!Operational Type Classes}
\emph{Operational type classes}~\cite{BS2011}
allow us to define a common notation 
for multiplication in any algebraic structure. 
First, we associate a class to the notion of \emph{multiplication} 
on any type $A$.

@@

\subsection{The Monoid Type Class}
\index{Coq!Type Classes}
We are now ready to  give a definition of the \texttt{Monoid} class, using
\texttt{*} as an infix operator in scope \coqscope{M} for the monoid  multiplication.

@@

\subsection{Monoids and Equivalence Relations}
\index{Coq!Generalized rewriting}
\index{Coq!Equivalence relations}

@@

\index{Coq!Type Classes}
\index{Coq!Proper}
\label{EMonoid-def}

@@

\index{Coq!Coercions}

@@

In order to make possible proof by rewriting on expressions which contain
the exponentiation operator, we have to prove that, whenever \texttt{$x$ == $y$},
the equality \texttt{$x^n$ == $y^n$} holds for any exponent \texttt{$n$}. 
For this purpose, we use the \texttt{Proper} class of module
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Classes.Morphisms.html}{Coq.Classes.Morphisms}
\index{Coq!Type Classes}
\index{Coq!Proper}
\begin{Coqsrc}
Global Instance power_proper :
     Proper (equiv ==> eq ==> equiv) power.
(* Proof omitted *)
\end{Coqsrc}

@@

\subsection{Correctness Proofs by Reflection}
\label{reflection-section}
\index{Coq!Proof by reflection}
Instead of letting the tactic \texttt{rewrite} look for contexts in which
setoid rewriting is possible, we propose to use (deterministic) computations for
obtaining a ``canonical'' form for terms generated from a variable \texttt{x}
by contructors associated with monoid multiplication and neutral element.

@@

\subsubsection{About parametricity}
\index{Coq!Parametricity}
Let us now present another tactic for proving chain correctness,
in the tradition of works on \emph{parametricity} and its use for 
proving properties on programs.
Strachey~\cite{Strachey:2000:FCP:609150.609208}
explores the nature of \emph{parametric
polymorphism}: ``\emph{Polymorphic functions behave uniformly for all types}''
then Reynolds~\cite{REYNOLDS83} formalizes this notion through binary relations.
Wadler~\cite{Wadler1989}, then Cohen \emph{et al.}~\cite{Cohen2013}
use this relation for deriving
 theorems about functions that operate on parametric
polymorphic types.

@@

\subsubsection{Definition:} 
\index{Coq!Type Classes}
\index{Coq!Proper}
A continuation \texttt{k:Fkont A} is \emph{proper}
if, whenever \linebreak[3] \texttt{x == y} holds, the computations \texttt{k x} and 
\texttt{k y} are equivalent.

@@

The following dependently-typed functions will help us to specify  formally
any correct chain generator.
\index{Coq!Dependently Typed Functions}

### 2021-01-15 doc/chapter-primrec.tex
comments and examples about the Ackermann part (to continue)

\subsection{Basic Libraries}

@@

\subsubsection{Functions of arbitrary arity}

@@

The magic of dependent types makes it possible to define recursively extensionnal equality between functions of the same arity.

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.extEqualNat.html}{Ackermann.extEqualNat}}
\begin{Coqsrc}
Fixpoint  extEqual (n : nat) : forall  (a b : naryFunc n), Prop :=
  match n with
    0 => fun a b => a = b
  | S p => fun a b => forall c, extEqual p (a c) (b c)
  end.
\end{Coqsrc}

@@

>>>>

\section{Basic Libraries}

@@

\subsection{Functions of arbitrary arity}

@@

The magic of dependent types makes it possible to define recursively extensional equality between functions of the same arity.

\index{Coq!Dependent types}
\index{Coq!Dependently typed functions}
\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.extEqualNat.html}{Ackermann.extEqualNat}}
\begin{Coqsrc}
Fixpoint  extEqual (n : nat) : forall  (a b : naryFunc n), Prop :=
  match n with
    0 => fun a b => a = b
  | S p => fun a b => forall c, extEqual p (a c) (b c)
  end.
\end{Coqsrc}

@@

\subsection{A Data-type for Primitive Recursive Functions}

The traditional definition of the set of primitive recursive functions is structured as an inductive definition 
in five rules: three base cases, and two construction rules. Primitive recursive functions are total functions from $\mathbb{N}^n$ to
$\mathbb{N}$, for some $n\in\mathbb{N}$.

\begin{description}
  \item[zero] the constant function of value $0$ is primitive recursive.
\item[S] The successor function $S:\mathbb{N}\rightarrow\mathbb{N}$ is primitive recursive.
 \item[projections] For any pair $0\leq i<n$, the projection $\pi_{n,i}: \mathbb{N}^n\rightarrow\mathbb{N}$, defined by $\pi_{n,i}(x_0,x_1,\dots,x_{n-1})=x_i$, is primitive recursive.
\item[composition] For any $n$ and $m$, if $h: \mathbb{N}^m\rightarrow\mathbb{N}$, and
$g_0,\dots, g_{m-1}$ are primitive recursive of $n$ arguments, then the function which maps any
tuple $(x_0,\dots,x_{n-1})$ to $h(g_0(x0,\dots,x_{n-1}),\dots, g_{m-1}(x0,\dots,x_{n-1}))$ is primitive recursive.
\item[primitive recursion]
If $g: \mathbb{N}^n\rightarrow\mathbb{N}$ and $h: \mathbb{N}^{n+2}\rightarrow\mathbb{N}$ are primitive recursive, then the function from $\mathbb{N}^{n+1}$ into $\mathbb{N}$ defined by
\begin{align}
f(0,x_1,\dots,x_n)&=g(x_1,\dots,x_n)\\
f(S(p),x_1,\dots,x_n)&=h(p,f(p, x_1,\dots,x_n),  x_1,\dots,x_n)
\end{align} 
is primitive recursive.
\end{description}

O'Connor's formalization of primitive recursive functions takes the form of two mutually inductive dependent data types, each constructor of which is associated with one of these  rules.
These two types are (\texttt{PrimRec $n$}) (primitive recursive functions of $n$ arguments), and
(\texttt{PrimRecs $n$ $m$}) ($m$-tuples of primitive recursive functions of $n$ arguments).

\index{Coq!Dependent types}
\index{Coq!Mutually inductive types}

\vspace{4pt}
\noindent
\emph{From \href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}}
\begin{Coqsrc}
Inductive PrimRec : nat -> Set :=
  | succFunc : PrimRec 1
  | zeroFunc : PrimRec 0
  | projFunc : forall n m : nat, m < n -> PrimRec n
  | composeFunc :
      forall (n m : nat) (g : PrimRecs n m) (h : PrimRec m), PrimRec n
  | primRecFunc :
      forall (n : nat) (g : PrimRec n) (h : PrimRec (S (S n))), 
      PrimRec (S n)
with PrimRecs : nat -> nat -> Set :=
  | PRnil : forall n : nat, PrimRecs n 0
  | PRcons : forall n m : nat, PrimRec n -> PrimRecs n m -> 
                   PrimRecs n (S m).
\end{Coqsrc}

\subsection{A little bit of Semantics} 
Please note that inhabitants of type (\texttt{PrimRec $n$}) are not \coq{} functions like \texttt{Nat.mul}, or factorial, etc. The data-type (\texttt{PrimRec $n$}) is indeed an abstract syntax for the language of primitive recursive functions. The bridge between this language and the word of usual functions
is an interpretation function (\texttt{evalprimRec $n$})  from  (\texttt{PrimRec $n$}) to  (\texttt{naryFunc $n$}) (with the helper (\texttt{evalprimRecS $n$ $m$})
from  (\texttt{PrimRecs $n$ $m$}) to  (\texttt{Vector.t (naryFunc $n$) $m$}).

Both functions are mutually defined through dependent pattern matching. We advise the readers who 
would feel uneasy with dependent types to consult Adam Chlipala's \emph{cpdt}  book~\cite{chlipalacpdt2011}. We leave it to the reader  to look at the helper functions in
\href{../theories/html/hydras.Ackermann.primRec.html}{Ackermann.primRec}.

\begin{Coqsrc}
Fixpoint evalPrimRec (n : nat) (f : PrimRec n) {struct f} : 
 naryFunc n :=
  match f in (PrimRec n) return (naryFunc n) with
  | succFunc => S
  | zeroFunc => 0
  | projFunc n m pf => evalProjFunc n m pf
  | composeFunc n m l f =>
      evalComposeFunc n m (evalPrimRecs _ _ l) (evalPrimRec _ f)
  | primRecFunc n g h =>
      evalPrimRecFunc n (evalPrimRec _ g) (evalPrimRec _ h)
  end

 with evalPrimRecs (n m : nat) (fs : PrimRecs n m) {struct fs} :
 Vector.t (naryFunc n) m :=
  match fs in (PrimRecs n m) return (Vector.t (naryFunc n) m) with
  | PRnil a => Vector.nil  (naryFunc a)
  | PRcons a b g gs =>
       Vector.cons _ (evalPrimRec _ g) _  (evalPrimRecs _ _ gs)
  end.
\end{Coqsrc}
Let us take as example the following term\footnote{Of course, we never typed this term \emph{verbatim}, but obtained as the result of some computation we leave it the reader to guess and reproduce.}:

\begin{Coqsrc}
Example weirdPR : PrimRec 1 :=
primRecFunc 0
  (composeFunc 0 1 (PRcons 0 0 zeroFunc (PRnil 0)) succFunc)
  (composeFunc 2 2
    (PRcons 2 1
      (composeFunc 2 1
         (PRcons 2 0 (projFunc 2 1 (le_n 2))
                 (PRnil 2))
         succFunc)
      (PRcons 2 0
        (composeFunc 2 1
          (PRcons 2 0
             (projFunc 2 0
                       (le_S 1 1 (le_n 1)))
             (PRnil 2))
          (projFunc 1 0 (le_n 1))) (PRnil 2)))
    (primRecFunc 1 (composeFunc 1 0 (PRnil 1) zeroFunc)
       (composeFunc 3 2
         (PRcons 3 1
            (projFunc 3 1 (le_S 2 2 (le_n 2)))
            (PRcons 3 0 (projFunc 3 0
                          (le_S 1 2
                                (le_S 1 1 (le_n 1))))
                    (PRnil 3)))
         (primRecFunc 1 (projFunc 1 0 (le_n 1))
                      (composeFunc 3 1
                          (PRcons 3 0
                                  (projFunc 3 1 (le_S 2 2 (le_n 2)))
                                  (PRnil 3))
                          succFunc))))). 
\end{Coqsrc}

Let us now interpret this term as an arithmetic function.

\begin{Coqsrc}
Example  mystery_fun : nat _> nat := evalPrimRec 1 weirdPR.

Compute List.map mystery_fun (0::1::2::3::4::5::6::nil) : list nat.
\end{Coqsrc}

\begin{Coqanswer}
 = 1 :: 1 :: 2 :: 6 :: 24 :: 120 :: 720 :: nil
     : list nat
\end{Coqanswer}

Ok, the term \texttt{weirdPR} looks to be a primitive recursive definition of the factorial function, although we haven't proved this fact yet. Fortunately, we will see in the following sections much simpler ways to prove that a given function is primitive recursive, whithout looking at an unreadable term.
